Splitting of the homology of the punctured mapping class group

### 2018-11-05 Introduction.tex
First draft ready, sent to cfb,palmer,kranhold,kiefner,soulie for comments

Let $\gg$ be the mapping class group of $\sg$, i.e. the group of isotopy classes of diffeomorphisms of $\sg$:
diffeomorphisms are required to fix $\partial\sg$ pointwise; similarly let $\ggm$ be the mapping class group of $\sgm$, i.e.
the group of isotopy classes of diffeomorphisms of $\sgm$ that fix $\partial\sgm$ pointwise and \emph{permute} the $m$ punctures.

Forgetting punctures yields a surjective map $\ggm\to\gg$ with kernel $\bms$,
the $m-$th \emph{braid group} of the surface $\sg$: we get the Birman exact sequence (see \cite{Birman:mcgbr}):
\begin{equation}
\label{eq:Birman}
1\to\bms\to\ggm\to\gg\to 1.
\end{equation}
% This short exact sequence corresponds to 
% \begin{equation}
%  \cms\to\mgm\to\mg.
% \end{equation}
% Here $\cms$ is the \emph{unordered} configuration space of $m$ points in the interior of $\Sigma$, i.e.
% \begin{equation}
%  \cms=\fms/\S_m=\set{(p_1,\dots,p_m)\in\pa{\mathring\sg}^m|p_i\neq p_j \;\forall i\neq j}/\S_m,
% \end{equation}
% the quotient of the \emph{ordered} configuration space of $m$ points in the interior of $\Sigma$ by the natural
% action of the symmetric group $\S_m$ that permutes the (ordered list of) points in the configuration; $\mg$ and $\mgm$
% are the moduli spaces of Riemann structures on $\sg$ and $\sgm$ respectively.

@@

Our main theorem is that the spectral sequence has already collapsed in its second page.
\begin{thm}
\label{thm:main}
For all $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
 H_l(\ggm;\Z_2)\cong \bigoplus_{k+q=l} H_k(\gg;H_q(\bms;\Z_2)).
\end{equation}
\end{thm}
Thus the computation of $H_*(\ggm;\Z_2)$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*(\bms;\Z_2)$: as we will see the $\ggm-$representation
$H_*(\bms;\Z_2)$ splits as a direct sum of symmetric powers of $H_1(\sg)$.
In particular this representation is symplectic: this had already benn stated in \cite{LM}.

The group $\gg$ is significantly smaller than $\ggm$, so the
computation of the right-hand side of equation \ref{eq:main}
should be more tractable.
WE DO THIS COMPUTATION FOR $g=1$.

>>>>

Let $\gg$ be the mapping class group of $\sg$, i.e. the group of isotopy classes of diffeomorphisms of $\sg$:
diffeomorphisms are required to fix $\partial\sg$ pointwise. Similarly let $\ggm$ be the mapping class group of $\sgm$, i.e.
the group of isotopy classes of diffeomorphisms of $\sgm$ that fix $\partial\sgm$ pointwise and \emph{permute} the $m$ punctures.

Forgetting the punctures gives a surjective map $\ggm\to\gg$ with kernel $\bms$,
the $m-$th \emph{braid group} of the surface $\sg$: we obtain the Birman exact sequence (see \cite{Birman:mcgbr}):
\begin{equation}
\label{eq:Birman}
1\to\bms\to\ggm\to\gg\to 1.
\end{equation}
% This short exact sequence corresponds to 
% \begin{equation}
%  \cms\to\mgm\to\mg.
% \end{equation}
% Here $\cms$ is the \emph{unordered} configuration space of $m$ points in the interior of $\Sigma$, i.e.
% \begin{equation}
%  \cms=\fms/\S_m=\set{(p_1,\dots,p_m)\in\pa{\mathring\sg}^m|p_i\neq p_j \;\forall i\neq j}/\S_m,
% \end{equation}
% the quotient of the \emph{ordered} configuration space of $m$ points in the interior of $\Sigma$ by the natural
% action of the symmetric group $\S_m$ that permutes the (ordered list of) points in the configuration; $\mg$ and $\mgm$
% are the moduli spaces of Riemann structures on $\sg$ and $\sgm$ respectively.

@@

The main theorem of this article states that this spectral sequence has already collapsed in its second page.
\begin{thm}
\label{thm:main}
For all $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
H_l(\ggm;\Z_2)\cong \bigoplus_{k+q=l} H_k(\gg;H_q(\bms;\Z_2)).
\end{equation}
\end{thm}
Thus the computation of $H_*(\ggm;\Z_2)$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*(\bms;\Z_2)$. We will see that the $\ggm-$representation
$H_*(\bms;\Z_2)$ splits as a direct sum of symmetric powers of $H_1(\sg)$ with symplectic action
(see theorem \ref{thm:Hbms*as*ggrep} in section \ref{sec:HBraidSurf}).

%In particular this representation is symplectic: this had already benn stated in \cite{LM}.

The group $\gg$ is significantly smaller than $\ggm$, so the
computation of the right-hand side of equation \ref{eq:main}
may be more tractable.
%WE DO THIS COMPUTATION FOR $g=1$.

### 2018-11-05 Preliminaries.tex
First draft ready, sent to cfb,palmer,kranhold,kiefner,soulie for comments

In this section we recollect some classical definitions and results about braid groups and mapping class groups.
By $\sg$ we will always denote a compact, smooth, orientable surface of genus $g$ with one parametrised
boundary component $\partial \sg$; by $\sgm$ we denote a surface $\sg$ with a choice of $m$ distinct
points in the interior: these points are called \emph{punctures} and are not assumed to be ordered
or labelled.

\begin{defn}
\label{defn:cms}
 The $m$-th \emph{ordered configuration space} of a surface $\sg$ is the space
\[
 F_m(\sg)=\set{(P_1,\dots,P_m) \in \pa{\mathring{\Sigma}_{g,1}}^{\times m}  \,|\,  P_i\neq P_j  \,\forall i\neq j}
\]
 Notice that we require the points of the configuration to lie in the interior of $\sg$.
 Thus $F_m(\sg)$ is naturally endowed with the structure of a smooth, orientable $2m-$dimensional
 manifold.

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the points of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m-$th unordered configuration space}
 of $\sg$ and denoted by $C_m(\sg)$; this space is also a $2m-$dimensional orientable manifold.

@@

@@

\begin{defn}
 \label{def:mcg}
 Let $\Diff(\sg;\partial\sg)$ be the group of diffeomorphisms of $\sg$ that fix $\partial\sg$ pointwise,
 endowed with the Whitney $C^{\infty}-$topology. We denote by $\gg=\pi_0(\Diff(\sg;\partial\sg))$ its group of connected
 components, called also the \emph{mapping class group} of $\sg$.

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and permute the $m$ punctures, and let $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$ be its
 group of connected components, called the mapping class group of $\sgm$.
 \end{defn}
 A classical result by Earle and Schatz (\cite{EarleSchatz}) ensures that the connected components
 of $\Diff(\sg)$ are contractible. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. a space of type $K(\gg,1)$; we call $\cF_{g,1}\to B\Diff(\sg;\partial\sg)$ the universal
 $\sg-$bundle
 \[
  \sg\times_{\Diff(\sg;\partial\sg)}E\Diff(\sg;\partial\sg)\to B\Diff(\sg;\partial\sg).
 \]
Applying the construction of the $m-$th unordered configuration space fiberwise we obtain a bundle
$C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg)$ with fiber $C_m(\sg)$.

The space $C_m(\cF_{g,1})$ is a classifying space for the
group $\ggm$ and the Birman exact sequence \ref{eq:Birman} is obtained by taking fundamental
groups of the spaces
\begin{equation}\label{eq:Birmanbundle}
C_m(\sg)\to C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg).
\end{equation}

@@

It will be useful, for many constructions, to choose an embedding
$\D\hookrightarrow\mrS$ near $\partial\S$ and to replace
$\Diff(\S;\partial\S)$ with its subgroup $\Diff(\S;\partial\S\cup\D)$ of diffeomorphisms
of $\S$ that fix pointwise both $\partial\S$ and $\D$: here when we say that $\D$ is embedded
\emph{near $\partial\S$} we mean that there is a compact subsurface $\S'\subset\S$
such that $\S$ is the boundary connected sum $\S'\natural\bar\D$.
From now on we suppose such an embedding to be fixed and we consider $\D$ as a subspace
of $\mrS$. In section \ref{sec:HBraidSurf} we will specify an
embedding in a particular model for $\mrS$ that we will define.

The construction of definition \ref{defn:cms} can be specialised to the open
surfaces $\D$ and $\S'$, yielding spaces $C_m(\D)$ and $C_m(\S')$ respectively.

The inclusion $\Diff(\S;\partial\S\cup\D)\subset\Diff(\S;\partial\S)$ is a homotopy
equivalence, hence the previous constructions are equivalent to the ones that we
give in the next definition.
\begin{defn}
\label{defn:universalSbundle}
We call $\cF_{\S,\D}\to B\Diff(\S;\partial\S\cup\D)$
the universal $\S-$bundle
 \[
  \S\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S\cup\D).
 \]
The group $\Diff(\S;\partial\S\cup\D)$ acts on $\S'$ by restriction of diffeomorphisms;
the space $\cF_{\S,\D}$ contains subspaces
\[
 \cF_{\S'}= \S'\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)
\]
and $\D\times B\Diff(\S;\partial\S\cup\D)$, fibering over $B\Diff(\S;\partial\S\cup\D)$
with fibers $\S'$ and $\D$ respectively.

Applying the construction of the $m-$th unordered configuration space fiberwise, we obtain spaces
$C_m(\cF_{\S,\D})$, $C_m(\cF_{\S'})$ and $C_m(\D)\times B\Diff(\S;\partial\S\cup\D)$,
all fibering over $B\Diff(\S;\partial\S\cup\D)$ fibers, respectively, $\cms$,
$C_m(\S')$ and $C_m(\D)$.

For all $0\leq p\leq m$ there is a natural map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which consists on the union of configurations:
 \[
  \mu(\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}})=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
We can apply this construction to each couple of fibers of $C_m(\cF_{\S,\D})$ and $C_{m-p}(\cF_{\S'})$
over the same point of $ B\Diff(\S;\partial\S\cup\D)$, obtaining a map $\mu^{\cF}$:
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
If we see the domain of $\mu^{\cF}$ as a fibered product
\[
 \pa{C_m(\D)\times B\Diff(\S;\partial\S\cup\D)}\times_{B\Diff(\S;\partial\S\cup\D)}C_{m-p}(\cF_{\S'}),
\]
then the map $\mu^{\cF}$ is a map of bundles over $B\Diff(\S;\partial\S\cup\D)$, and corresponding
map on fibers is precisely $\mu$.
\end{defn}
The fiber bundle \ref{eq:Birmanbundle} corresponding to the Birman exact sequence \ref{eq:Birman}
can now be replaced by the following one
\begin{equation}\label{eq:BirmanbundleD}
\cms\to C_m(\cF_{\S,\D})\to B\Diff(\S;\partial\S\cup\D).
\end{equation}
We conclude this section by recalling the structure of $H_*(C_m(\D))$.

The cohomology of $C_m(\D)$ with coefficients in $\Z_2$ was first computed by Fuchs in
\cite{Fuchs:CohomBraidModtwo}: we will recall this computation more in detail in section
\ref{sec:HBraidSurf}, where we will make in detail a similar computation.

In \cite[Chap.III]{CLM} Cohen considers $\coprod_{m\geq 0}C_m(\D)$ as an
algebra over the operad of little $2-$cubes, and in particular describes its
homology with coefficients in $\Z_2$ as follows:
\begin{equation}
 \label{eq:Cohen}
H_*\pa{\coprod_{m\geq 0}C_m(\D)}\simeq \Z_2\left[Q^j\epsilon \, |\, j\geq 0\right].
\end{equation}

Here $\epsilon\in H_0(C_1(\D))$ is the fundamental class, and for all
$k,m\geq 0$ we denote by $Q\colon H_k(C_m(\D))\to H_{2k+1}(C_{2m}(\D))$
the first Dyer-Lashof operation. In particular $Q^j\epsilon$ is the generator of
$H_{2^j-1}(C_{2^j}(\D))\simeq\Z_2$.

The isomorphism in equation \ref{eq:Cohen} is
an isomorphism of bigraded rings: the left-hand side is a ring with the Pontryagin product,
and the right-hand side is a polynomial ring in the infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$;
the bidegree is given by
$*$, called \emph{degree}, which is the homological degree,
and $m$, called \emph{weight}, which is the index of the connected component
on which the homology class is supported, or informally the number of points
involved in the construction of a homology class.

In this article we will only need the isomorphism in equation \ref{eq:Cohen} to hold as
an isomorphism of bigraded $\Z_2-$vector spaces.
In particular for all choices of natural numbers $(\alpha_j)_{j\geq 0}$ with all but finitely
many $\alpha_j=0$, we will consider the monomial $\prod_j\pa{Q^j\epsilon}^{\alpha_j}$, which is
a homology class in some bidegree $(*,m)$: all that
we are going to need is that the set of these monomials is a bigraded basis for the
left-hand side of equation \ref{eq:Cohen}.

>>>>

In this section we recollect some classical definitions and results about braid groups and mapping class groups.
Let $\sg$ be a compact, smooth, orientable surface of genus $g$ with one parametrised
boundary component $\partial \sg$. Let $\sgm$ be $\sg$ with a choice of $m$ distinct
points in the interior: these points are called \emph{punctures} and are not assumed to be ordered.

\begin{defn}
\label{defn:cms}
 The $m$-th \emph{ordered configuration space} of a surface $\sg$ is the space
\[
 F_m(\sg)=\set{(P_1,\dots,P_m) \in \pa{\mathring{\Sigma}_{g,1}}^{\times m}  \,|\,  P_i\neq P_j  \;\forall i\neq j}.
\]

 Notice that we require the points of the configuration to lie in the interior of $\sg$;
 $F_m(\sg)$ is a smooth, orientable $2m-$dimensional manifold.

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the points of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m-$th unordered configuration space}
 of $\sg$ and is denoted by $C_m(\sg)$; this space is also a $2m-$dimensional orientable manifold
 (see figure \ref{fig:unordered}).

@@

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/unordered.png}
 \caption{A configuration in the space $C_5(\Sigma_{1,1})$}
\label{fig:unordered}
\end{figure}

@@

\begin{defn}
 \label{def:mcg}
 Let $\Diff(\sg;\partial\sg)$ be the group of diffeomorphisms of $\sg$ that fix $\partial\sg$ pointwise,
 endowed with the Whitney $C^{\infty}-$topology. We denote by $\gg=\pi_0(\Diff(\sg;\partial\sg))$ its group of connected
 components, which is called the \emph{mapping class group} of $\sg$.

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and restrict to a permutation of the $m$ punctures, and let $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$ be its
 group of connected components, called the mapping class group of $\sgm$.
 \end{defn}
 A classical result by Earle and Schatz (\cite{EarleSchatz}) ensures that the connected components
 of $\Diff(\sg)$ are contractible. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. a space of type $K(\gg,1)$; we call $\cF_{g,1}\to B\Diff(\sg;\partial\sg)$ the universal
 $\sg-$bundle
 \[
  \cF_{g,1}\colon=\sg\times_{\Diff(\sg;\partial\sg)}E\Diff(\sg;\partial\sg)\to B\Diff(\sg;\partial\sg).
 \]
Applying the construction of the $m-$th unordered configuration space fiberwise we obtain a bundle
$C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg)$ with fiber $C_m(\sg)$.

The space $C_m(\cF_{g,1})$ is a classifying space for the
group $\ggm$ and the Birman exact sequence \ref{eq:Birman} is obtained by taking fundamental
groups of the aspherical spaces
\begin{equation}\label{eq:Birmanbundle}
C_m(\sg)\to C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg).
\end{equation}

@@

It will be useful, for many constructions, to choose an embedding
$\D\hookrightarrow\mrS$ near $\partial\S$ and to replace
$\Diff(\S;\partial\S)$ with its subgroup $\Diff(\S;\partial\S\cup\D)$ of diffeomorphisms
of $\S$ that fix pointwise both $\partial\S$ and $\D$. Saying that $\D$ is embedded
\emph{near $\partial\S$} means that there is a compact subsurface $\S'\subset\S$
such that $\S$ is the boundary connected sum $\S'\natural\bar\D$.
From now on we suppose such an embedding to be fixed and we consider $\D$ as a subspace
of $\mrS$ (see figure \ref{fig:SS'D}). In section \ref{sec:HBraidSurf}, definition \ref{defn:Tsg},
we will specify an embedding $\D\hookrightarrow\mrS$ using a particular model for $\mrS$.

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/SS'D.png}
 \caption{A splitting of $\S$ as $\S'\natural\bar\D$.}
\label{fig:SS'D}
\end{figure}

The construction of definition \ref{defn:cms} can be specialised to the
surfaces $\D$ and $\S'$, yielding spaces $C_m(\D)$ and $C_m(\S')$ respectively.

The inclusion $\Diff(\S;\partial\S\cup\D)\subset\Diff(\S;\partial\S)$ is a homotopy
equivalence, hence also the induced map
\[
B\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S)
\]
is a homotopy equivalence. We will replace the previous construction with the following,
equivalent ones.
\begin{defn}
\label{defn:universalSbundle}
We call $\cF_{\S,\D}\to B\Diff(\S;\partial\S\cup\D)$
the universal $\S-$bundle
 \[
  \cF_{\S,\D}\colon=\S\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S\cup\D).
 \]
Notice that $\Diff(\S;\partial\S\cup\D)$ acts on $\S'$ by restriction of diffeomorphisms, and acts trivially
on $\D$; therefore $\cF_{\S,\D}$ contains subspaces
\[
 \cF_{\S'}= \S'\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)
\]
and $\D\times B\Diff(\S;\partial\S\cup\D)$; these subspaces fiber over $B\Diff(\S;\partial\S\cup\D)$
with fibers $\S'$ and $\D$ respectively.

Applying the construction of the $m-$th unordered configuration space fiberwise, we obtain spaces
$C_m(\cF_{\S,\D})$, $C_m(\cF_{\S'})$ and $C_m(\D)\times B\Diff(\S;\partial\S\cup\D)$,
all fibering over $B\Diff(\S;\partial\S\cup\D)$ with fibers, respectively, $\cms$,
$C_m(\S')$ and $C_m(\D)$.
\end{defn}

\begin{defn}
For all $0\leq p\leq m$ there is a natural map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which takes the union of configurations (see figure \ref{fig:defmu}):
 \[
  \mu\pa{\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}}}=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
\begin{figure}\centering
 \includegraphics[scale=0.7]{figures/defmu.png}
 \caption{The product of two configurations in $C_2(\D)$ and $C_4(\S)$.}
\label{fig:defmu}
\end{figure}

We can apply this construction to each couple of fibers of $C_m(\cF_{\S,\D})$ and $C_{m-p}(\cF_{\S'})$
over the same point of $ B\Diff(\S;\partial\S\cup\D)$, obtaining a map $\mu^{\cF}$:
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
If we see the domain of $\mu^{\cF}$ as a fibered product of bundles over $B\Diff(\S;\partial\S\cup\D)$
\[
 \pa{C_m(\D)\times B\Diff(\S;\partial\S\cup\D)}\times_{B\Diff(\S;\partial\S\cup\D)}C_{m-p}(\cF_{\S'}),
\]
then the map $\mu^{\cF}$ is also a map of bundles over $B\Diff(\S;\partial\S\cup\D)$, and the corresponding
map on fibers is precisely $\mu$.
\end{defn}
The fiber bundle \ref{eq:Birmanbundle} corresponding to the Birman exact sequence \ref{eq:Birman}
can now be replaced by the following one
\begin{equation}\label{eq:BirmanbundleD}
\cms\to C_m(\cF_{\S,\D})\to B\Diff(\S;\partial\S\cup\D).
\end{equation}
We conclude this section by recalling the structure of $H_*(C_m(\D))$.

The cohomology of $C_m(\D)$ with coefficients in $\Z_2$ was first computed by Fuchs in
\cite{Fuchs:CohomBraidModtwo}: we will recall this computation in section
\ref{sec:HBraidSurf}, where we will make a similar computation in detail.

In \cite[Chap.III]{CLM} Cohen considers the space $\coprod_{m\geq 0}C_m(\D)$ as an
algebra over the operad of little $2-$cubes, and describes its
$\Z_2-$homology as follows:
\begin{equation}
 \label{eq:Cohen}
H_*\pa{\coprod_{m\geq 0}C_m(\D)}\simeq \Z_2\left[Q^j\epsilon \, |\, j\geq 0\right].
\end{equation}

Here $\epsilon\in H_0(C_1(\D))$ is the fundamental class, and for all
$k,m\geq 0$ we denote by $Q\colon H_k(C_m(\D))\to H_{2k+1}(C_{2m}(\D))$
the first Dyer-Lashof operation. In particular $Q^j\epsilon$ is the generator of
$H_{2^j-1}(C_{2^j}(\D))\simeq\Z_2$. See figure \ref{fig:monomial}

The isomorphism in equation \ref{eq:Cohen} is
an isomorphism of bigraded rings. The left-hand side is a ring with the Pontryagin product
coming from the action of the little $2-$cubes,
and the right-hand side is a polynomial ring in infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$.
The bidegree is given by the homological degree
$*$, that we simply call the \emph{degree},
and by the index $m$ of the connected component
on which the homology class is supported (informally, the number of points
involved in the construction of a homology class), that we call the \emph{weight}.

 \begin{figure}\centering
 \includegraphics[scale=0.7]{figures/monomial.png}
 \caption{The class in $H_5(C_9(\D))$ corresponding to the monomial $\epsilon\cdot(Q\epsilon)^2\cdot Q^2\epsilon$.}
\label{fig:monomial}
\end{figure}

In this article we will only need the isomorphism in equation \ref{eq:Cohen} to hold as
an isomorphism of bigraded $\Z_2-$vector spaces.
In particular for all choices of natural numbers $(\alpha_j)_{j\geq 0}$ with all but finitely
many $\alpha_j=0$, we will consider the monomial $\prod_j\pa{Q^j\epsilon}^{\alpha_j}$, corresponding to
a homology class in some bidegree $(*,m)$: all that
we are going to need is that the set of these monomials is a bigraded basis for the
left-hand side of equation \ref{eq:Cohen}.

### 2018-11-05 Rational.tex
First draft ready, sent to cfb,palmer,kranhold,kiefner,soulie for comments

\section{A rational counterexample}
In this section we prove that a statement as in theorem \ref{thm:Hbms*as*ggrep}
cannot hold if we consider homology with coefficients in $\Q$.

We recall that $H_*(\cms;\Q)$ has been computed \emph{as a bigraded $\Q-$vector space}
in \cite{BCM} and, more recently, in \cite{Knudson}; a description of it as a
$\gg-$representation seems to be still missing in the literature.

In particular we prove the following fact.
\begin{thm}
 \label{thm:counterexample}
 Let $g\geq 2$ and $m\geq 2$; then $H_2(C_2(\S);\Q)$ is not a symplectic
 representation of $\gg$.
\end{thm}
\begin{proof}
 We will use the following strategy:
 \begin{itemize}
  \item we define a homology class $[a]\in H_2(C_2(\S);\Q)$ represented by a cycle $a$;
  \item we prove that $[a]\neq 0$ by computing the algebraic intersection of
  $[a]$ with a homology class
  in $\tH_2(\cms^c;\Q)$; here it is worth stressing that for all $m$ the manifold
  $\cms$ is orientable: therefore we can apply Poincaré-Lefschetz duality
  also with rational coefficients;
  \item we define another homology class $[b]\in H_2(\cms;\Q)$ and show that
  $[b]$ is mapped to $[b]+2[a]$ by some element in the Torelli group $\mathcal{I}_{g,1}$.
 \end{itemize}
Recall that the Torelli group $\mathcal{I}_{g,1}$ is the kernel of the surjective homomorphism
$\gg\to Sp_{2g}(\Z)$; if an element of the Torelli group acts non-trivially on
some class in $H_2(C_2(\S);\Q)$, then this is not a symplectic representation of $\gg$.

@@

Let $\tup=(0,(u_i,v_i))$ with $u_1=2$ and all other $u_i$'s, as well as all $v_i$'s, equal to zero;
then the map
\[
\phi_{\tup}\colon (\Delta^{\tup},\partial\Delta^{\tup})\to (C_2(\S)^c,\infty)
\]
maps the fundamental class in $H_2(\Delta^{\tup},\partial\Delta^{\tup};\Q)$ to a homology
class in $H_2(C_2(\S);\Q)$; if we see $e^{\tup}$ as a properly embedded, orientable submanifold
of $C_2(\S)$, then we are just considering its fundamental class $[e^{\tup}]\in H_2(C_2(\S)^c)$.

The cell $e^{\tup}$ intersects once, transversely the torus $a$, therefore $[a]\cap[e^{\tup}]=\pm 1$,
where the signs depends on how we have chosen orientations on $a$, $e^{\tup}$ and $C_2(\S)$ itself;
in particular $[a]\neq 0$.

The action of $\gg$ on isotopy classes of simple closed curves is transitive, so we can repeat
the construction of the torus $a$ with any other copy of parallel, non-separating simple closed curves
$c$ and $c'$, and the resulting class $[a]$ will always be non-zero in $H_2(C_2(\S);\Q)$.

Let now $d$ and $d'$ be disjoint, non-separating simple closed curves such that
cutting $\S$ along $d$ and $d'$ we obtain a subsurface $\S'\simeq\Sigma_{1,2}$ of genus 1, with boundary
components $d$ and $d'$ (here we need that the genus of $\S$ is at least 2);
suppose moreover that $c$ is a non-separating simple closed curve in $\S'$,
and let $c',c''$ be two parallel copies of $c$ in $\Sigma_{1,2}$, one on each side of a small tubular
neighborhood of $c$.
Then $d,d',c',c''$ are the boundary of a subsurface $\S''\simeq\Sigma_{0,4}\subset\S'$.
Orient all curves $d,d',c,c',c''$ in such a way that the following equalities hold:
\begin{itemize}
 \item $[d]=[d']\in H_1(\S;\Q)$, as witnessed by the subsurface $\S'$;
 \item $[c]=[c']=[c'']\in H_1(\S;\Q)$;
 \item $[d]-[d']+[c']-[c'']=0\in H_1(\S\setminus c;\Q)$, as witnessed by the subsurface $\S''$.
\end{itemize}
See picture PICTURE.

@@

Consider now an element of the Torelli group that fixes $c$ and maps $d$ to $d'$ preserving
the orientation, for example the bounding pair represented in picture PICTURE:
then the class $b$ is mapped to $b'=b+2a\neq b$.
\end{proof}

>>>>

\section{A rational counterexample}
In this section we prove that a statement as in theorem \ref{thm:Hbms*as*ggrep}
cannot hold if we consider homology with coefficients in $\Q$. We still don't know
if the analogue of theorem \ref{thm:main} holds in homology with coefficients in $\Q$
or in fields of odd characteristic.

We recall that $H_*(C_m(\Sigma_{g,1});\Q)$ has been computed \emph{as a bigraded $\Q-$vector space}
by B\"odigheimer, Cohen and Milgram in \cite{BCM}, and more recently by Knudson in \cite{Knudson}. A description of
these homology groups as a
$\gg-$representation seems to be still missing in the literature.

We will prove the following theorem:
\begin{thm}
 \label{thm:counterexample}
 Let $g\geq 2$ and $m\geq 2$; then $H_2(C_2(\Sigma_{g,1});\Q)$ is not a symplectic
 representation of $\gg$.
\end{thm}
\begin{proof}
 Let again $\S=\Sigma_{g,1}$. We will use the following strategy:
 \begin{itemize}
  \item we define a homology class $[a]\in H_2(C_2(\S);\Q)$ represented by a cycle $a$;
  \item we prove that $[a]\neq 0$ by computing the algebraic intersection of
  $[a]$ with a homology class
  in $\tH_2(\cms^c;\Q)$; here it is worth stressing that the manifold
  $\cms$ is orientable: therefore we can apply Poincaré-Lefschetz duality
  also with rational coefficients;
  \item we define another homology class $[b]\in H_2(\cms;\Q)$ and show that
  $[b]$ is mapped to $[b]+2[a]$ by some element in the Torelli group $\mathcal{I}_{g,1}$.
 \end{itemize}
Recall that the Torelli group $\mathcal{I}_{g,1}$ is the kernel of the surjective homomorphism
$\gg\to Sp_{2g}(\Z)$; if an element of the Torelli group acts non-trivially on
some class in $H_2(C_2(\S);\Q)$, then this is not a symplectic representation of $\gg$.

@@

Let $\tup=(0,(u_i,v_i))$ with $u_1=2$ and all other $u_i$'s, as well as all $v_i$'s, equal to zero.
Then the map
\[
\phi_{\tup}\colon (\Delta^{\tup},\partial\Delta^{\tup})\to (C_2(\S)^c,\infty)
\]
maps the fundamental class in $H_2(\Delta^{\tup},\partial\Delta^{\tup};\Q)$ to a homology
class in $\tH_2(C_2(\S)^c;\Q)$.
% ; if we see $e^{\tup}$ as a properly embedded, orientable submanifold
% of $C_2(\S)$, then we are just considering its fundamental class $[e^{\tup}]\in H_2(C_2(\S)^c)$.

The cell $e^{\tup}$ intersects once, transversely the torus $a$, therefore the
algebraic intersection $[a]\cap[e^{\tup}]$ is $\pm 1$,
where the signs depends on how we have chosen orientations on $a$, $e^{\tup}$ and $C_2(\S)$ itself;
in particular $[a]\neq 0$.

The action of $\gg$ on isotopy classes of simple closed curves is transitive, so we can repeat
the construction of the torus $a$ with any other copy of parallel, non-separating simple closed curves
$c$ and $c'$, and the resulting class $[a]$ will always be non-trivial in $H_2(C_2(\S);\Q)$.

See figure \ref{fig:rational} to visualize the following discussion.
Let $d$ and $d'$ be disjoint, non-separating simple closed curves such that
cutting $\S$ along $d$ and $d'$ we obtain a subsurface $\S'\simeq\Sigma_{1,2}$ of genus 1, with boundary
components $d$ and $d'$ (here we need that the genus of $\S$ is at least 2);
suppose moreover that $c$ is a non-separating simple closed curve in $\S'$,
and let $c',c''$ be two parallel copies of $c$ in $\Sigma_{1,2}$, one on each side of a small tubular
neighborhood of $c$.
Then $d,d',c',c''$ are the boundary of a subsurface $\S''\simeq\Sigma_{0,4}\subset\S'$.
Orient all curves $d,d',c,c',c''$ in such a way that the following equalities hold:
\begin{itemize}
 \item $[d]=[d']\in H_1(\S;\Q)$, as witnessed by the subsurface $\S'$;
 \item $[c]=[c']=[c'']\in H_1(\S;\Q)$;
 \item $[d]-[d']+[c']-[c'']=0\in H_1(\S\setminus c;\Q)$, as witnessed by the subsurface $\S''$.
\end{itemize}

\begin{figure}\centering
 \includegraphics[scale=1.0]{figures/rational.png}
 \caption{The curves $d,d',c,c',c'',\gamma,\gamma''$ and the subsurfaces $\S',\S''$.}
\label{fig:rational}
\end{figure}

@@

Consider now an element of the Torelli group that fixes $c$ and maps $d$ to $d'$ preserving
the orientation, for example the bounding pair $D_{\gamma}\circ D_{\gamma'}^{-1}$, where
$\gamma$ and $\gamma'$ are represented in figure \ref{fig:rational}:
then the class $b$ is mapped to $b'=b+2a\neq b$.
\end{proof}

The previous proof works word by word if we replace $\Q$ by any field of odd characteristic. Moreover
all the arguments in the previous proof can be adapted to show that $H_*(F_2(\S);\mathbb{F})$ is
not a symplectic representation of $\gg$, where $F_2(\S)$ is the ordered configuration space
and $\mathbb{F}$ is \emph{any} field, including $\Z_2$. The result of theorem \ref{thm:Hbms*as*ggrep}
is peculiar of unordered configurations and characteristic 2.
